\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{circuitikz}

\title{Qubit Introduction}
\author{Martina Paola Calvi}
\date{\today}

\begin{document}

\maketitle
\tableofcontents
\newpage

\section{Introduction to Qubits}
\subsection{Computers and Computation}

\subsubsection*{Models of Classical Computation}
Classical computation can be implemented through various physical models:
\begin{itemize}
    \item \textbf{Mechanical}: Curta calculator, Digi-comp I, Babbage's difference engine
    \item \textbf{Electrical}: Electronic circuits and transistor-based systems
    \item \textbf{Optical}: Photonic computing systems
    \item \textbf{Biological}: DNA and molecular computing
\end{itemize}

\subsubsection*{Conceptual Models}
Key theoretical frameworks for computation include:
\begin{itemize}
    \item Turing Machine (mathematical model of computation)
    \item Cellular Automata
    \item Von Neumann architecture (CPU, Memory, Bus, I/O)
    \item Logic-in-Memory computing paradigm
\end{itemize}

\subsubsection*{Quantum Computing Development}

In the early 1980s, Richard Feynman proposed that simulating quantum systems requires quantum computers. This led to the fundamental question: what algorithms could provide a quantum advantage for real-world problems?

Two major milestones occurred in the mid-1990s:
\begin{enumerate}
    \item Discovery of Shor's algorithm for factoring (addressing an important practical problem)
    \item Development of quantum error-correcting codes by Shor and colleagues
\end{enumerate}

Since then, research has focused on both the underlying physics and hardware development for quantum computers. Today, operating quantum computers with more than 100 qubits exist, with IBM's Condor exceeding 1000 qubits.

\subsection{Data Representation: Classical vs. Quantum}

Classical computers use bits (0 or 1), while quantum computers use quantum bits (qubits) that can exist in superposition: $|\psi\rangle = \alpha|0\rangle + \beta|1\rangle$ where $|\alpha|^2 + |\beta|^2 = 1$.

\subsubsection*{Quantum Parallelism}

Unlike classical bits, quantum systems exhibit superposition, allowing exponential encoding of information. While $N$ classical bits represent a single $N$-bit state, $N$ qubits exist in a superposition of all $2^N$ possible states simultaneously.

For example, with $N=3$ qubits:
$$|\psi\rangle = \sum_{i=0}^{7} c_i|i\rangle$$

where each $c_i$ is a complex amplitude and $\sum_{i=0}^{7}|c_i|^2 = 1$.

This quantum parallelism enables quantum algorithms to process all $2^N$ computational paths concurrently, providing potential exponential speedup over classical approaches. However, measurement collapses this superposition to a single state, requiring careful algorithm design to extract useful information.

\subsection{DiVincenzo Criteria for Quantum Computing}
A viable qubit must satisfy the DiVincenzo Criteria:
\begin{enumerate}
    \item Scalability of the physical system
    \item Ability to initialize qubit states
    \item Long coherence times ($T_1$, $T_2$, $T_\phi$)
    \item Universal quantum gate operations
    \item High-fidelity quantum operations
    \item Interconversion between stationary and flying qubits
    \item Faithful transmission of flying qubits
\end{enumerate}

\subsection{Qubit Figure of Merits}

\subsubsection*{Qubit Robustness and Coherence Time}
Qubits lose information through interaction with the environment via:
\begin{itemize}
    \item Energy relaxation ($T_1$): loss of energy to the environment
    \item Dephasing ($T_\phi$): loss of phase coherence
\end{itemize}

The effective coherence time is $\frac{1}{T_2} = \frac{1}{T_\phi} + \frac{1}{2T_1}$, representing the average time a qubit remains coherent.

\subsubsection*{Gate Speed and Figure of Merit}
The number of gates performable within qubit lifetime determines quantum advantage: $N_{\text{gates}} \sim t_{\text{gate}}/T_2$.

\subsubsection*{Gate Fidelity}
Gate fidelity quantifies how accurately quantum gate operations are performed, measuring the closeness between ideal and actual operations under imperfections and noise. High-fidelity operations are crucial for reliable quantum computation and can be determined through Process Tomography or Randomized Benchmarking.

\subsection{Bits vs. Qubits}
\subsubsection*{Classical vs. Quantum Gates}

Classical computers use Boolean logic gates (such as NOT and AND) to form a universal set of operations. Quantum computers similarly use quantum gates, but with fundamentally different properties:

\begin{itemize}
    \item Classical gates: irreversible, map multiple inputs to outputs
    \item Quantum gates: unitary and reversible, preserve quantum information
    \item Single quantum gates (e.g., X-gate) can manipulate qubit states
    \item Universal quantum computation requires a combination of single and two-qubit gates
\end{itemize}

The X-gate example shows practical application: applying an X-gate to a qubit in state $|0\rangle$ transitions it to $|1\rangle$, achieved experimentally through a $\pi$-pulse envelope.

\subsubsection*{Circuits in Space vs. Circuits in Time}

Classical logic operates as \textbf{circuits in space}:
\begin{itemize}
    \item Example: NOT gate maps $0 \to 1$ and $1 \to 0$
    \item Horizontal axis represents space
    \item Input and output exist at different physical locations
    \item Measurements can be performed simultaneously on different wires
\end{itemize}

In contrast, quantum logic operates as \textbf{circuits in time}:
\begin{itemize}
    \item Example: X gate (quantum NOT) acts on qubit states $|0\rangle$ and $|1\rangle$
    \item Horizontal axis represents time
    \item Input and output correspond to the same qubit before and after the operation
    \item State is sequentially updated through time evolution
\end{itemize}

\textbf{Key distinction:}
\begin{itemize}
    \item Classical gates connect wires in space, allowing parallel independent operations
    \item Quantum gates are unitary operations that evolve the state of qubits in time
    \item Each gate application modifies and overwrites the qubit's state
\end{itemize}
\subsubsection*{Two-Qubit Gates}

Two-qubit gates enable quantum entanglement and are essential for universal quantum computation.

\paragraph{CNOT Gate (Controlled-NOT):}

The CNOT gate operates on two qubits: a control qubit and a target qubit. It applies an X gate to the target if the control qubit is $|1\rangle$.

$$\text{CNOT} = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \end{pmatrix}$$

Truth table:

\begin{center}
\begin{tabular}{cc|cc}
\hline
Control & Target & Control & Target \\
\hline
$|0\rangle$ & $|0\rangle$ & $|0\rangle$ & $|0\rangle$ \\
$|0\rangle$ & $|1\rangle$ & $|0\rangle$ & $|1\rangle$ \\
$|1\rangle$ & $|0\rangle$ & $|1\rangle$ & $|1\rangle$ \\
$|1\rangle$ & $|1\rangle$ & $|1\rangle$ & $|0\rangle$ \\
\hline
\end{tabular}
\end{center}

For a control qubit in superposition $|\psi\rangle_A = \frac{1}{\sqrt{2}}(|0\rangle + |1\rangle)$ and target in state $|0\rangle_B$:

$$|\psi\rangle_{\text{out}} = \text{CNOT}(|\psi\rangle_A \otimes |0\rangle_B) = \frac{1}{\sqrt{2}}(|0\rangle_A|0\rangle_B + |1\rangle_A|1\rangle_B)$$

This creates an entangled Bell state, where the qubits are correlated regardless of measurement basis.

\subsection{Qubit States}

\subsubsection*{General Qubit State}

Quantum mechanics tells us that any qubit system can exist in a superposition of states. The general state of a quantum bit is described by:
$$|\psi\rangle = \alpha|0\rangle + \beta|1\rangle$$

where $\alpha$ and $\beta$ are complex numbers, and the normalization constraint $\langle\psi|\psi\rangle = 1$ requires that:
$$|\alpha|^2 + |\beta|^2 = 1$$

The states $|0\rangle$ and $|1\rangle$ represent the computational basis as two-dimensional vectors:
$$|0\rangle = \begin{pmatrix} 1 \\ 0 \end{pmatrix}, \quad |1\rangle = \begin{pmatrix} 0 \\ 1 \end{pmatrix}$$

A general superposition state can be expressed as a weighted sum of basis states:
$$|\psi\rangle = \alpha|0\rangle + \beta|1\rangle = \alpha\begin{pmatrix} 1 \\ 0 \end{pmatrix} + \beta\begin{pmatrix} 0 \\ 1 \end{pmatrix} = \begin{pmatrix} \alpha \\ \beta \end{pmatrix}$$

\subsubsection*{Bra-Ket Notation and Inner Products}

The transpose complex conjugate of a ket $|\psi\rangle$ is denoted as a bra:
$$|\psi\rangle^\dagger = \langle\psi| = \begin{pmatrix} \alpha^* & \beta^* \end{pmatrix}$$

where the $\dagger$ superscript indicates the Hermitian conjugate (transpose complex conjugate).

The inner product forms the bra-ket:
$$\langle\psi|\psi\rangle = |\alpha|^2 + |\beta|^2 = 1$$

This represents the normalization condition ensuring the total probability is conserved.

\subsubsection*{Measurement of Qubits}

When a qubit in superposition $|\psi\rangle = \alpha|0\rangle + \beta|1\rangle$ is measured in the computational basis:

\begin{itemize}
    \item The measurement is an active process where the apparatus interacts with the qubit
    \item The qubit state collapses to either $|0\rangle$ or $|1\rangle$
    \item The probability of measuring $|0\rangle$ is $P(0) = |\alpha|^2$
    \item The probability of measuring $|1\rangle$ is $P(1) = |\beta|^2$
    \item After measurement, the qubit is no longer in superposition
    \item Only partial information is extracted—you cannot access the full quantum state
    \item Information obtained depends on the chosen measurement basis
\end{itemize}

This measurement in the computational basis $\{|0\rangle, |1\rangle\}$ is called \textbf{Z-measurement}. Other measurement bases exist (X-basis, Y-basis), but the computational basis is most common in quantum computing systems.

\subsubsection*{The Born Rule}
The probability of measuring state $|a\rangle$ from $|\psi\rangle$ is:
$$P(a) = |\langle a|\psi\rangle|^2$$

Common measurement bases include:
\begin{itemize}
    \item Computational (Z-basis): $\{|0\rangle, |1\rangle\}$
    \item X-basis: $\{|+\rangle, |-\rangle\}$ where $|+\rangle = \frac{1}{\sqrt{2}}(|0\rangle + |1\rangle)$
    \item Y-basis: $\{|+i\rangle, |-i\rangle\}$ where $|+i\rangle = \frac{1}{\sqrt{2}}(|0\rangle + i|1\rangle)$
\end{itemize}

\subsubsection*{TODO:Bloch Sphere Representation}

Any single-qubit pure state can be represented on the Bloch sphere:
$$|\psi\rangle = \cos\frac{\theta}{2}|0\rangle + e^{i\phi}\sin\frac{\theta}{2}|1\rangle$$

where $\theta \in [0, \pi]$ and $\phi \in [0, 2\pi]$.

The Bloch vector is $\vec{r} = (\sin\theta\cos\phi, \sin\theta\sin\phi, \cos\theta)$.

Key states: $|0\rangle \to (0,0,1)$, $|1\rangle \to (0,0,-1)$, $|+\rangle \to (1,0,0)$, $|-\rangle \to (-1,0,0)$.

\subsection{Single-Qubit Gates}

Single-qubit gates are unitary transformations that evolve qubit states. A unitary matrix $U$ satisfies $U^\dagger U = UU^\dagger = I$.

\subsubsection*{Pauli Gates}

\paragraph{Pauli-X (NOT gate):}
$$\sigma_x = \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}, \quad \sigma_x|0\rangle = |1\rangle, \quad \sigma_x|1\rangle = |0\rangle$$

Represents a $\pi$ rotation around the x-axis.

\paragraph{Pauli-Z gate:}
$$\sigma_z = \begin{pmatrix} 1 & 0 \\ 0 & -1 \end{pmatrix}, \quad \sigma_z|0\rangle = |0\rangle, \quad \sigma_z|1\rangle = -|1\rangle$$

Represents a $\pi$ rotation around the z-axis.

\paragraph{Pauli-Y gate:}
$$\sigma_y = \begin{pmatrix} 0 & -i \\ i & 0 \end{pmatrix} = -i\sigma_x\sigma_z$$

Represents a $\pi$ rotation around the y-axis. All Pauli matrices satisfy $\sigma_i^2 = I$ and are Hermitian.

\subsubsection*{Commutation Relations}

$$[\sigma_x, \sigma_y] = 2i\sigma_z, \quad [\sigma_y, \sigma_z] = 2i\sigma_x, \quad [\sigma_z, \sigma_x] = 2i\sigma_y$$

\subsubsection*{Anticommutation Relations}

$$\{\sigma_x, \sigma_y\} = 0, \quad \{\sigma_y, \sigma_z\} = 0, \quad \{\sigma_z, \sigma_x\} = 0$$

\subsubsection*{Rotation Gates}

General rotation around axis $n$ by angle $\theta$:
$$R_n(\theta) = e^{-i\theta\sigma_n/2} = \cos\frac{\theta}{2}I - i\sin\frac{\theta}{2}\sigma_n$$

\subsubsection*{Hadamard Gate}

$$H = \frac{1}{\sqrt{2}}\begin{pmatrix} 1 & 1 \\ 1 & -1 \end{pmatrix}$$

Creates superposition: $H|0\rangle = |+\rangle$, $H|1\rangle = |-\rangle$.

\subsubsection*{Phase Gate (S-gate)}

$$S = \begin{pmatrix} 1 & 0 \\ 0 & i \end{pmatrix}$$

Adds a $90°$ phase shift: $S|+\rangle = |+i\rangle$, $S|-\rangle = |-i\rangle$.

\subsection{Multiple Gate Applications}

Sequential gate applications are computed via matrix multiplication (right to left in mathematical notation, left to right in circuits).

\subsection{Density Matrix Formalism}

Real quantum systems cannot be perfectly isolated and are described by the density matrix $\hat{\rho}$:

\subsubsection*{Pure States}
$$\hat{\rho} = |\psi\rangle\langle\psi|$$

\subsubsection*{Mixed States (Ensemble)}
$$\hat{\rho} = \sum_i p_i |\psi_i\rangle\langle\psi_i|$$

where $\sum_i p_i = 1$.

\subsubsection*{Properties}
\begin{itemize}
    \item Hermitian: $\hat{\rho} = \hat{\rho}^\dagger$
    \item Trace normalized: $\text{tr}(\hat{\rho}) = 1$
    \item 2×2 matrix form: $\hat{\rho} = \begin{pmatrix} \rho_{11} & \rho_{12} \\ \rho_{21} & \rho_{22} \end{pmatrix}$
    \item Diagonal elements represent probabilities
    \item Off-diagonal elements represent coherences (with $\rho_{12} = \rho_{21}^*$)
\end{itemize}

Pure vs. mixed states can be distinguished by $\text{tr}(\hat{\rho}^2) = 1$ (pure) or $\text{tr}(\hat{\rho}^2) < 1$ (mixed).
\newpage
\section{Superconducting Qubits}
\subsubsection*{Superconductivity Review}

In 1911, Heike Kamerlingh Onnes discovered superconductivity by observing that the resistivity of mercury vanished abruptly at 4.2 K. In 1933, Walther Meissner and Robert Ochsenfeld discovered that below a critical temperature $T_C$, a superconductor expels magnetic fields from its interior—the \textbf{Meissner effect}—demonstrating that superconductivity is a distinct thermodynamic state, not merely a perfect conductor.

In 1957, John Bardeen, Leon Cooper, and J. Robert Schrieffer published the microscopic \textbf{BCS Theory} of superconductivity. The key insight is that electrons form \textbf{Cooper pairs}—correlated pairs with opposite momentum and spin. These pairs form a coherent \textbf{Bose-Einstein condensate} described by a single macroscopic wave function:
$$\psi(\mathbf{r}, t) = \sqrt{\rho_s(\mathbf{r}, t)} e^{i\phi(\mathbf{r}, t)}$$

where $\rho_s$ is the density of Cooper pairs and $\phi$ is the macroscopic phase. The superconducting current is related to the phase gradient:
$$J_s = \frac{n_s e \hbar}{m} \nabla\phi - \frac{n_s e^2}{m c} \mathbf{A}$$

This phase rigidity—where variations in $\phi$ cost energy—is the microscopic origin of superconductivity.

\subsubsection*{The Josephson Junction}

In 1962, Brian D. Josephson analyzed a junction of two superconductors separated by a thin insulating barrier (0.1–1 nm). Despite the barrier, Cooper pairs tunnel across via the quantum tunneling effect. The junction is characterized by two fundamental relations:

\paragraph{DC (Current-Phase) Relation:}
$$I = I_C \sin \phi$$

where $I_C$ is the critical current (maximum supercurrent without dissipation) and $\phi = \phi_2 - \phi_1$ is the phase difference.

\paragraph{AC (Voltage-Phase) Relation:}
$$V = \frac{\Phi_0}{2\pi}\frac{d\phi}{dt}$$

where $\Phi_0 = h/2e = 2\pi\hbar/2e$ is the \textbf{magnetic flux quantum}. This shows that voltage drives phase evolution in time.

\subsubsection*{Josephson Inductance and Energy}

The junction exhibits non-linear inductance. Differentiating the current-phase relation:
$$\frac{dI}{dt} = I_C \cos(\phi) \frac{d\phi}{dt}$$

Substituting $\frac{d\phi}{dt} = \frac{2\pi V}{\Phi_0}$ yields:
$$L_J(\phi) = \frac{\Phi_0}{2\pi I_C \cos(\phi)} = \frac{L_{J0}}{\cos(\phi)}, \quad L_{J0} = \frac{\Phi_0}{2\pi I_C}$$

The Josephson energy (potential energy stored in phase) is:
$$E_J(\phi) = -E_J \cos \phi$$

where $E_J = \frac{\Phi_0 I_C}{2\pi} = \frac{\hbar I_C}{2e}$ is the \textbf{Josephson coupling energy}.

\subsubsection*{Energy Scales: $E_J$ and $E_C$}

Superconducting qubits are characterized by two competing energy scales:

\begin{itemize}
    \item \textbf{Josephson Energy ($E_J$):} Tunneling energy of Cooper pairs across the barrier.
    \item \textbf{Charging Energy ($E_C$):} Energy cost to add one Cooper pair to the island: $E_C = \frac{e^2}{2C}$.
\end{itemize}

The ratio $E_J / E_C$ determines the qubit regime and dynamical properties.

\subsubsection*{Circuit Quantization}

Using canonical quantization, flux $\Phi$ and charge $Q$ become non-commuting operators:
$$[\hat{\Phi}, \hat{Q}] = i\hbar$$

The Hamiltonian for a superconducting qubit with a Josephson junction replacing the inductor is:
$$\hat{H} = \frac{\hat{Q}^2}{2C} - E_J \cos\left(\frac{2\pi \hat{\Phi}}{\Phi_0}\right)$$

In terms of the phase operator $\hat{\phi} = 2\pi\hat{\Phi}/\Phi_0$ and the Cooper pair number operator $\hat{n} = \hat{Q}/2e$:
$$\hat{H} = 4E_C(\hat{n} - n_g)^2 - E_J \cos \hat{\phi}$$

where $n_g$ is the gate charge offset. The term $(\hat{n} - n_g)^2$ means the qubit is sensitive to charge noise from the environment.

\subsubsection*{The Transmon Regime}

The \textbf{Transmon} (Transmission-line shunted plasma oscillation qubit), introduced by Koch et al. (2007), is the industry standard. It operates in the regime $E_J / E_C \gg 1$ (typically 20–100) to minimize charge noise sensitivity.

\paragraph{Exponential Noise Suppression:}
By increasing $E_J / E_C$, energy level dispersions (charge noise sensitivity) are exponentially suppressed:
$$\text{Dispersion} \propto \exp\left(-\sqrt{\frac{8 E_J}{E_C}}\right)$$

For $E_J / E_C = 50$: $\exp(-\sqrt{400}) \approx \exp(-20) \approx 10^{-9}$. This provides dephasing times $T_2 > 100~\mu\text{s}$.

\paragraph{Controlled Anharmonicity:}
Despite increased immunity to noise, the Transmon retains small anharmonicity. Expanding the cosine potential:
$$-E_J \cos \hat{\phi} \approx E_J\left(\frac{\hat{\phi}^2}{2} - \frac{\hat{\phi}^4}{24} + \cdots\right)$$

In terms of ladder operators and applying the Rotating Wave Approximation:
$$\hat{H} \approx \hbar\omega_T \hat{b}^\dagger\hat{b} - \frac{E_C}{12}(\hat{b} + \hat{b}^\dagger)^4$$

where $\omega_T = \sqrt{8 E_J E_C}/\hbar - E_C/\hbar$ is the transition frequency. The anharmonicity is:
$$\alpha = E_{1\to 2} - E_{0\to 1} = -E_C$$

Typical values: $E_J \sim 20$ GHz, $E_C \sim 300\text{--}400$ MHz, yielding anharmonicity of 6–8% and gate times $T_\pi \approx 4\text{--}20$ ns.

\paragraph{Typical Performance Metrics:}
\begin{itemize}
    \item Relaxation time: $T_1 \sim 200\text{--}500~\mu\text{s}$
    \item Dephasing time: $T_2 \sim 100\text{--}300~\mu\text{s}$
    \item Gate fidelity: $> 99\%$
    \item Readout fidelity: $> 95\%$
\end{itemize}

\subsubsection*{Other Superconducting Qubit Types}

Beyond the Transmon, alternative architectures include:

\begin{itemize}
    \item \textbf{Cooper Pair Box (CPB):} The original charge qubit (Nakamura et al., Nature 1999), operating in the regime $E_J < E_C$. Highly sensitive to charge noise; historically important but superseded by Transmon.
    
    \item \textbf{Flux Qubit:} A superconducting loop with 3 Josephson junctions, operated near half a flux quantum ($\Phi_{\text{ext}} \approx \Phi_0/2$). The two persistent current states form the qubit basis. Sensitive to flux noise but with intermediate coherence times.
    
    \item \textbf{Fluxonium:} An island connected to ground through a superinductance (many large junctions in parallel) with a small junction providing strong non-linearity. High anharmonicity ($\alpha \sim 1$ GHz), robust against both charge and flux noise. Coherence times $T_2 > 1$ ms have been demonstrated.
    
    \item \textbf{Phase Qubit:} Exploits energy quantization in a double-well potential formed by a junction operated near critical current. Historically important (UCSB, 2002–2007) but coherence times are limited.
\end{itemize}

\subsubsection*{Decoherence Mechanisms and Coherence Times}

Decoherence is characterized by two fundamental timescales:

\begin{itemize}
    \item \textbf{$T_1$ (Relaxation):} Energy decay from $|1\rangle$ to $|0\rangle$. Sources: dielectric losses in the substrate, dielectric loss at the tunnel junction, Purcell damping from environmental modes, quasiparticle poisoning. State-of-the-art Transmons achieve $T_1 \sim 200\text{--}500~\mu\text{s}$.
    
    \item \textbf{$T_2$ (Dephasing):} Coherence decay. Related to $T_1$ by $T_2^{-1} = (2T_1)^{-1} + T_\phi^{-1}$, where $T_\phi$ is pure dephasing caused by 1/f flux noise, surface charge defects, and other low-frequency noise. Best Transmons achieve $T_2 \sim 100\text{--}300~\mu\text{s}$.
\end{itemize}


\newpage
\section{Circuit QED}

\subsection{From Atomic Physics to Quantum Optics}

The foundation of Circuit QED lies in coherent light-matter interaction at the quantum level. In atomic physics, discrete energy levels allow photons to drive transitions between states. The simplest useful model reduces an atom to a two-level system (TLS) with ground state $|0\rangle$ and excited state $|1\rangle$.

\subsubsection*{The Two-Level System}

The transition energy is:
$$E_{01} = E_1 - E_0 = \hbar\omega_{01}$$

Coupling to a classical oscillating electric field is described by:
$$H = -\vec{d} \cdot \vec{E}(t), \quad E(t) = E_0 \cos(\omega_{01} t)$$

where $d$ is the transition dipole moment. Near-resonant driving produces Rabi oscillations—coherent energy exchange between atom and field.

\subsubsection*{Vacuum Fluctuations and Field Confinement}

A key question motivates Cavity QED: can vacuum fluctuations alone produce significant effects on a qubit? The vacuum electric field amplitude in a single-mode cavity is:
$$E_{\text{zpf}} = \sqrt{\frac{\hbar\omega_r}{2\varepsilon_0 V}}$$

In optical cavities, $V \sim \lambda^3$ gives $E_{\text{zpf}} \sim 10^{-4}$ V/m. Superconducting circuits achieve $V_{\text{eff}} \sim 10^{-3}\lambda^3$, yielding $E_{\text{zpf}} \approx 0.2$ V/m—orders of magnitude larger. This tight confinement enables single-photon effects.

\subsection{Cavity QED and the Fabry-Pérot Resonator}

Traditional Cavity QED places atoms inside Fabry-Pérot cavities—two highly reflective mirrors separated by distance $L$. Resonant modes satisfy:
$$2L = \lambda_m m, \quad f_m = \frac{mc}{2L}$$

where $m$ is the mode number. The quality factor quantifies resonator performance:
$$Q = \frac{f_{\text{resonance}}}{\Delta f_{3\text{dB}}} = \frac{\omega_r}{\kappa}$$

Superconducting microwave resonators achieve $Q \sim 10^4$--$10^6$, corresponding to photon lifetimes $1/\kappa$ of microseconds.

The vacuum Rabi coupling constant is:
$$g = \frac{d \cdot E_{\text{zpf}}}{\hbar}$$

In optical cavity QED: $g/2\pi \sim 50$ kHz. In circuit QED: $g/2\pi \sim 0$--$1$ GHz—five orders of magnitude stronger due to large artificial dipole moments and tight field confinement.

\subsection{Circuit QED: Artificial Atoms and Coplanar Waveguides}

Circuit QED transplants cavity QED physics to a microchip, replacing atoms with superconducting qubits and Fabry-Pérot cavities with coplanar waveguide (CPW) resonators.

\subsubsection*{Advantages of Superconducting Qubits}

Superconducting qubits possess effective dipole moments several hundred to thousand times larger than atoms because charge is distributed over micrometer-scale distances:

\begin{itemize}
    \item Large engineered dipole moment $\to$ strong coupling $g$ (hundreds of MHz)
    \item On-chip frequencies determined by design (lithography), not fixed by nature
    \item Integrated fabrication $\to$ scalable multi-qubit processors
    \item Fast gate operations (nanosecond timescale) enabled by large $g$
\end{itemize}

\subsubsection*{Coplanar Waveguide Resonators}

The CPW resonator consists of a superconducting transmission line (Nb or Al on sapphire) with center conductor separated from ground planes. At microwave frequencies (3--8 GHz), the resonator length is $L \sim \lambda/2$. The electromagnetic field is confined to volume:
$$V_{\text{eff}} \sim w \cdot d \cdot L \sim 10^{-3}\lambda^3$$

where $w$ and $d$ are conductor width and gap (sub-micron). This yields zero-point voltage fluctuations:
$$V_{\text{zpf}} = \sqrt{\frac{\hbar\omega_r}{2C_r}} \approx 1~\mu\text{V}$$

The enormous field concentration enables coupling rates in the hundreds of MHz range.

\subsection{Resonator Quantization and One-Mode Approximation}

\subsubsection*{Quantum LC Oscillator}

A CPW section is modeled as a lumped LC circuit with resonance frequency:
$$\omega_r \approx \frac{1}{\sqrt{L_r C_r}}$$

Quantizing via canonical commutation relations $[\hat{Q}, \hat{\Phi}] = i\hbar$ and introducing ladder operators:
$$\hat{\Phi} = \Phi_{\text{zpf}}(\hat{a} + \hat{a}^\dagger), \quad \hat{Q} = iQ_{\text{zpf}}(\hat{a}^\dagger - \hat{a})$$

where $\Phi_{\text{zpf}} = \sqrt{\hbar/(2\omega_r C_r)}$, the Hamiltonian becomes:
$$\hat{H}_r = \hbar\omega_r \hat{a}^\dagger \hat{a}$$

\subsubsection*{One-Mode Approximation}

A real transmission line supports infinitely many modes at frequencies $\omega_1, \omega_2, \omega_3, \ldots$ with spacing of several GHz. The linewidth $\kappa_m \sim$ MHz is much smaller than mode spacing:
$$\kappa_m \ll |\omega_{m+1} - \omega_m|$$

Near resonance, the qubit couples appreciably to only one mode $\omega_r$. All others decouple, justifying the single-mode approximation.

\subsection{The Full Circuit QED Hamiltonian}

The total system Hamiltonian combines resonator, qubit, and interaction. Using the two-level approximation for the transmon and ladder operators:
$$\hat{H} = \hbar\omega_r \hat{a}^\dagger \hat{a} + \frac{\hbar\omega_{01}}{2}\hat{\sigma}_z - g(\hat{\sigma}_+ - \hat{\sigma}_-)(\hat{a}^\dagger - \hat{a})$$

where $\hat{a}$ ($\hat{\sigma}_\pm$) are resonator (qubit) operators, and the coupling strength is:
$$g = \omega_r \frac{C_g}{C}\left(\frac{E_J}{2E_C}\right)^{1/4}\sqrt{\frac{\pi Z_r}{R_k}}$$

with $Z_r \approx 50~\Omega$ (resonator impedance) and $R_k = h/e^2 \approx 25.8~\text{k}\Omega$ (von Klitzing constant).

\subsection{The Jaynes-Cummings Model}

The Jaynes-Cummings (JC) Hamiltonian is the cornerstone of cavity and circuit QED. It emerges from the full Hamiltonian via the Rotating Wave Approximation.

\subsubsection*{Rotating Wave Approximation}

Expanding the interaction term yields four products:
$$-g(\hat{\sigma}_+ \hat{a}^\dagger - \hat{\sigma}_+ \hat{a} - \hat{\sigma}_- \hat{a}^\dagger + \hat{\sigma}_- \hat{a})$$

The terms $\hat{\sigma}_+ \hat{a}^\dagger$ and $\hat{\sigma}_- \hat{a}$ represent energy non-conserving processes oscillating at $\omega_r + \omega_{01}$. Near resonance ($\omega_r \approx \omega_{01}$), these oscillate rapidly and can be dropped. Energy-conserving terms $\hat{\sigma}_+ \hat{a}$ and $\hat{\sigma}_- \hat{a}^\dagger$ evolve slowly.

After RWA, the Jaynes-Cummings Hamiltonian is:
$$\hat{H}_{\text{JC}} = \hbar\omega_r \hat{a}^\dagger \hat{a} + \frac{\hbar\omega_{01}}{2}\hat{\sigma}_z + \hbar g(\hat{a}^\dagger \hat{\sigma}_- + \hat{a}\hat{\sigma}_+)$$

The interaction terms enable coherent photon-qubit exchange while conserving energy.

\subsubsection*{Dressed States and Vacuum Rabi Splitting}

Exactly at resonance ($\Delta = \omega_{01} - \omega_r = 0$), the JC model yields energy eigenstates:
$$|+\rangle = \frac{|1,0\rangle + |0,1\rangle}{\sqrt{2}}, \quad |-\rangle = \frac{|1,0\rangle - |0,1\rangle}{\sqrt{2}}$$

with energies $E_\pm = \hbar\omega_r \pm \hbar g$. The energy splitting $2\hbar g$ at resonance is the vacuum Rabi splitting, observable in transmission spectroscopy as two peaks—direct evidence of strong light-matter coupling.

\subsection{Qubit Readout: The Dispersive Regime}

Resonant coupling enables fast energy exchange (useful for gates) but is destructive for non-demolition measurement. Readout uses the dispersive regime.

\subsubsection*{Dispersive Hamiltonian}

When the qubit is far detuned from the resonator ($\Delta = \omega_{01} - \omega_r \gg g, \kappa$), second-order perturbation theory yields:
$$\hat{H}_{\text{disp}} \approx \hbar(\omega_r + \chi\hat{\sigma}_z)\hat{a}^\dagger \hat{a} + \frac{\hbar\omega_{01}}{2}\hat{\sigma}_z$$

where the dispersive shift is:
$$\chi = \frac{g^2}{\Delta}$$

The resonator frequency becomes qubit-state-dependent:
\begin{itemize}
    \item Qubit state $|0\rangle \to \omega_r - \chi$
    \item Qubit state $|1\rangle \to \omega_r + \chi$
\end{itemize}

\subsubsection*{Quantum Non-Demolition Readout}

By sending a probe tone at frequency $\omega_{\text{probe}} \approx \omega_r$, the reflected (or transmitted) signal acquires a state-dependent phase shift:
$$\phi(|0\rangle) = -\arctan\left(\frac{2(\omega_{\text{probe}} - \omega_r + \chi)}{\kappa}\right)$$
$$\phi(|1\rangle) = -\arctan\left(\frac{2(\omega_{\text{probe}} - \omega_r - \chi)}{\kappa}\right)$$

Measuring $\Delta\phi = \phi(|0\rangle) - \phi(|1\rangle)$ determines the qubit state without exchanging energy. Readout fidelity is limited only by noise and integration time.

\subsubsection*{Typical Circuit QED Parameters}

\begin{itemize}
    \item Coupling: $g/2\pi \approx 50$--$300$ MHz
    \item Detuning: $\Delta/2\pi \approx 0.5$--$3$ GHz (ensuring $\Delta \gg g$)
    \item Dispersive shift: $\chi/2\pi = g^2/\Delta \approx 0.5$--$10$ MHz
    \item 2D resonator linewidth: $\kappa/2\pi \approx 5$--$10$ MHz
    \item 3D cavity linewidth: $\kappa/2\pi \approx 1$--$10$ kHz
\end{itemize}

\subsection{Single-Qubit Control}

\subsubsection*{XY Control (Rabi Driving)}

Microwave pulses at frequency $\omega_d \approx \omega_{01}$ drive Rabi oscillations. In the rotating frame (frame rotating at $\omega_d$), the effective Hamiltonian is:
$$\hat{H}_{\text{drive}} = \frac{\hbar\Delta_d}{2}\hat{\sigma}_z + \frac{\hbar\Omega_R}{2}(\cos\phi \hat{\sigma}_x + \sin\phi \hat{\sigma}_y)$$

where $\Delta_d = \omega_{01} - \omega_d$ is the drive detuning and $\Omega_R = d \cdot E_0 / \hbar$ is the Rabi frequency controlled by pulse amplitude $E_0$. The phase $\phi$ determines the rotation axis (X or Y gate).

In the dispersive regime, control photons are off-resonance with the cavity and mostly reflected. Careful pulse shaping (DRAG—Derivative Removal via Adiabatic Gate) suppresses unwanted qubit state transitions.

\subsubsection*{Z Control (Frequency Tuning)}

Modern transmons incorporate a SQUID loop (Superconducting Quantum Interference Device) replacing the single Josephson junction. The effective Josephson energy becomes:
$$E_J(\Phi_{\text{ext}}) = E_{J,\text{max}} \left|\cos\left(\pi \frac{\Phi_{\text{ext}}}{\Phi_0}\right)\right|$$

where $\Phi_{\text{ext}}$ is external magnetic flux through the SQUID and $\Phi_0 = h/(2e)$ is the flux quantum. Since $\omega_{01} \propto \sqrt{E_J}$, tuning $\Phi_{\text{ext}}$ via a local Z-control line enables fast (nanosecond) qubit frequency control.

\subsection{The Transmon and Xmon Qubits}

\subsubsection*{Transmon: Charge Insensitivity}

The Cooper Pair Box has Hamiltonian:
$$H_{\text{CPB}} = 4E_C(\hat{n} - n_g)^2 - E_J \cos\hat{\phi}$$

where $n_g$ is the gate charge offset. The CPB is highly charge-sensitive, making it susceptible to charge noise.

The transmon regime ($E_J/E_C \gg 1$, typically 50--100) is achieved by adding a large shunt capacitor. Results:
\begin{itemize}
    \item Energy levels become nearly charge-insensitive (exponential suppression: $e^{-\sqrt{8E_J/E_C}}$)
    \item Qubit frequency: $\omega_{01} \approx \sqrt{8E_C E_J}/\hbar - E_C/\hbar$
    \item Anharmonicity: $\alpha = (E_{12} - E_{01})/\hbar \approx -E_C/\hbar$ (typically $-100$ to $-300$ MHz)
    \item Coherence times: $T_1 \sim 200$--$500~\mu\text{s}$, $T_2 \sim 100$--$300~\mu\text{s}$
    \item Gate fidelity: $> 99\%$
\end{itemize}

\subsubsection*{Xmon: Cross-Shaped Capacitor}

The Xmon qubit replaces the transmon's capacitor with a cross-shaped ('X') metal pad. Advantages:
\begin{itemize}
    \item Large shunt capacitance in compact footprint
    \item Simplified multi-qubit layout (four arms connect to readout, XY drive, Z control, quantum bus)
    \item Reduced parasitic coupling between qubits (geometric symmetry)
    \item Identical Hamiltonian to transmon
    \item Achieves $T_1, T_2 \sim 10$s--$100$s of microseconds
\end{itemize}

\subsubsection*{Meander Resonators}

Modern processors employ meandering (serpentine) CPW resonators, where the transmission line is folded to achieve resonance at shorter physical footprint. Benefits:
\begin{itemize}
    \item Reduced chip area and crosstalk
    \item Simplified multi-qubit layout
    \item Same resonance frequency as straight resonators
\end{itemize}

\subsection{Scalable Multi-Qubit Processors}

Circuit QED provides a complete platform for scalable quantum processors:

\begin{center}
\begin{tabular}{lll}
\hline
\textbf{Component} & \textbf{Function} & \textbf{Implementation} \\
\hline
Readout Resonator & Measure qubit state & Meandered CPW \\
Xmon Island & Store quantum state & Al cross capacitor \\
SQUID/JJ & Non-linearity \& tunability & Al/$\text{Al}_2\text{O}_3$/Al junction \\
XY Control Line & Rotate qubits (gates) & Microwave drive line \\
Z Control Line & Tune qubit frequency & Local flux line \\
Quantum Bus & Entangle neighboring qubits & CPW coupler \\
\hline
\end{tabular}
\end{center}

These elements enable coherent control via resonant microwave drives and high-fidelity readout via dispersive interactions, forming the foundation of superconducting quantum processors with 10s to 100s of qubits.

\newpage
\section{Qubit Control}
\subsubsection*{Recap on Readout}

All superconducting qubits are currently measured primarily in the circuit quantum electrodynamics architecture. The circuit QED architecture involves a qubit coupled to a resonator, which in turn is coupled to a feed line, simulating the atom-cavity interaction of cavity electrodynamics.

In the dispersive regime, when the qubit is far detuned from the resonator frequency compared with their coupling rate $g$ and the resonator linewidth $\kappa$ (so $\Delta \gg g, \kappa$), the Jaynes-Cummings Hamiltonian can be approximated as:

$$H \approx (\omega_r + \chi \hat{\sigma}_z) \hat{a}^\dagger \hat{a} + \frac{\omega_{01}}{2} \hat{\sigma}_z$$

The resonator acquires a qubit-state-dependent shift. The hybridization of the qubit-resonator states results in decay of the qubit into the environment via the resonator's loss-rate $\kappa$, called \textbf{Purcell decay}:

$$\gamma_p = \frac{g^2}{\Delta^2} \kappa$$

This puts a constraint on how high $\kappa$ can be. The ratio $\chi / \kappa$ is typically set to unity for optimal readout (balancing dispersive shift and cavity linewidth).

\subsubsection*{Purcell Filters}

In some designs, \textbf{Purcell filters} are incorporated—frequency-selective elements placed between the resonator and transmission line. They act:
\begin{itemize}
    \item Transparent at $\omega_r$ (measurement photons can leak)
    \item Opaque at $\omega_q$ (qubit-frequency photons are confined)
\end{itemize}

This design decouples measurement speed from qubit lifetime: $\kappa_{\text{eff}}(\omega_q) \ll \kappa(\omega_r)$.

\subsection{Controlling the Qubit: The Bloch Sphere Picture}

Reading out the qubit state is only half the task. To run quantum algorithms, we must also prepare arbitrary qubit states and perform single-qubit gates—unitary rotations of the state vector on the Bloch sphere.

For a transmon qubit, there are two physically distinct control mechanisms:

\begin{itemize}
    \item \textbf{Electric (Capacitive / XY) Control:} A resonant microwave drive rotates the Bloch vector around an axis in the equatorial plane. The rotation axis is set by the phase of the drive, and the rotation angle by the pulse area.
    \item \textbf{Magnetic (Flux / Z) Control:} Tuning the qubit frequency via an external magnetic flux through the SQUID loop rotates the Bloch vector around the Z-axis.
\end{itemize}

\subsection{Capacitive (XY) Control: Derivation}

\subsubsection*{Circuit Model}

A microwave drive voltage $V_d(t)$ is connected to the qubit island via a small coupling capacitor $C_d$ (with $C_d \ll C_\Sigma$, where $C_\Sigma = C + C_d$ is the total qubit capacitance). The coupling is deliberately weak to preserve qubit coherence: a small $C_d$ reduces the admittance seen from the drive line and limits energy leakage into the $50~\Omega$ wiring impedance $R_w$.

The full circuit Hamiltonian (qubit plus drive coupling) is:
$$H = \frac{\hat{Q}^2}{2C_\Sigma} + \frac{\hat{\Phi}^2}{2L} + \frac{C_d}{C_\Sigma} V_d(t) \hat{Q}$$

The first two terms are the qubit oscillator. The third is the drive coupling: it is linear in the charge operator $\hat{Q}$ because the drive voltage couples electrostatically to the qubit island.

\subsubsection*{Quantizing the Charge Operator}

Expressing the charge operator in terms of ladder operators (as done for the LC oscillator):
$$\hat{Q} = -i Q_{\text{zpf}} (\hat{a} - \hat{a}^\dagger), \quad Q_{\text{zpf}} = \sqrt{\frac{\hbar}{2Z_0}}, \quad Z_0 = \sqrt{\frac{L}{C_\Sigma}}$$

where $Q_{\text{zpf}}$ is the charge zero-point fluctuation and $Z_0$ is the characteristic impedance of the qubit oscillator. Substituting into the drive coupling term:
$$H_{\text{drive}} = \frac{C_d}{C_\Sigma} V_d(t) \cdot (-i Q_{\text{zpf}})(\hat{a} - \hat{a}^\dagger)$$

\subsubsection*{Two-Level Truncation}

For the transmon in the qubit subspace $\{|0\rangle, |1\rangle\}$, the ladder operators map to Pauli raising/lowering operators: $\hat{a} \to \hat{\sigma}_-$ and $\hat{a}^\dagger \to \hat{\sigma}_+$. The relevant Pauli matrix identities are:
$$\hat{\sigma}_+ = |1\rangle\langle 0|, \quad \hat{\sigma}_- = |0\rangle\langle 1|$$
$$\hat{\sigma}_x = \hat{\sigma}_+ + \hat{\sigma}_-, \quad \hat{\sigma}_y = i(\hat{\sigma}_- - \hat{\sigma}_+)$$

After truncation, $\hat{a} - \hat{a}^\dagger \to \hat{\sigma}_- - \hat{\sigma}_+ = i\hat{\sigma}_y$. The full Hamiltonian in the two-level approximation becomes:
$$\hat{H} = -\frac{\hbar\omega_q}{2} \hat{\sigma}_z + \Omega V_d(t) \hat{\sigma}_y$$

where $\omega_q = (E_1 - E_0)/\hbar$ is the qubit transition frequency and $\Omega = \frac{C_d}{C_\Sigma}\sqrt{\frac{\hbar}{2Z_0}}$ is the coupling strength (units: V$^{-1}$).

\subsection{The Rotating Frame}

The static qubit Hamiltonian $H_0 = -\frac{\hbar\omega_q}{2}\hat{\sigma}_z$ causes the Bloch vector to precess around the Z-axis at frequency $\omega_q \sim 5$ GHz. This rapid precession makes it difficult to visualize and analyze slow drive-induced rotations. The solution is to move into the rotating frame—a frame co-rotating with the qubit at $\omega_q$.

\subsubsection*{Unitary Transformation}

The rotating-frame transformation is defined by the unitary operator:
$$U(t) = e^{i\omega_q t \hat{\sigma}_z / 2}$$

This represents a rotation of angle $\omega_q t$ around the Z-axis of the Bloch sphere. The new state in the rotating frame is:
$$|\psi_{\text{rf}}\rangle = U(t)|\psi\rangle$$

The effective Hamiltonian governing the evolution of $|\psi_{\text{rf}}\rangle$ is found via:
$$\hat{H}' = U(t) \hat{H} U^\dagger(t) - i\hbar U(t) \frac{dU^\dagger(t)}{dt}$$

The second term, $-i\hbar U(t)\frac{dU^\dagger}{dt} = -\frac{\hbar\omega_q}{2}\hat{\sigma}_z$, exactly cancels the free evolution term $U \hat{H}_0 U^\dagger = -\frac{\hbar\omega_q}{2}\hat{\sigma}_z$. This is the defining property of the rotating frame: the free precession disappears and only the drive-induced dynamics remain.

\subsubsection*{Transforming the Drive Term}

The drive term $\hat{H}_d = \Omega V_d(t) \hat{\sigma}_y$ must also be transformed. The key identity, derived from the Baker-Campbell-Hausdorff formula, is:
$$U(t) \hat{\sigma}_y U^\dagger(t) = \cos(\omega_q t) \hat{\sigma}_y - \sin(\omega_q t) \hat{\sigma}_x$$

Physically: $\hat{\sigma}_y$ in the lab frame becomes a combination of $\hat{\sigma}_x$ and $\hat{\sigma}_y$ in the rotating frame, rotating at $\omega_q$. The drive Hamiltonian in the rotating frame is therefore:
$$\hat{H}'_d = \Omega V_d(t) [\cos(\omega_q t) \hat{\sigma}_y - \sin(\omega_q t) \hat{\sigma}_x]$$

\subsection{I/Q Decomposition and the Rotating Wave Approximation}

\subsubsection*{I/Q Signal Decomposition}

A general monochromatic drive voltage at frequency $\omega_d$ with phase $\phi$ and envelope $s(t)$ can be written as:
$$V_d(t) = V_0 s(t) [\cos(\phi) \sin(\omega_d t) + \sin(\phi) \cos(\omega_d t)]$$

This is the standard I/Q (in-phase / quadrature) decomposition:
$$V_d(t) = V_0 s(t) [I \sin(\omega_d t) + Q \cos(\omega_d t)], \quad I = \cos(\phi), \quad Q = \sin(\phi)$$

Substituting the I/Q form into the rotating-frame drive Hamiltonian and expanding all products of trigonometric functions yields terms oscillating at the difference frequency $\delta\omega = \omega_q - \omega_d$ and the sum frequency $\omega_q + \omega_d$.

\subsubsection*{Rotating Wave Approximation (RWA)}

When the drive is near resonance ($\omega_d \approx \omega_q$), the sum-frequency terms oscillate at $\sim 2\omega_q \sim 10$ GHz, which is much faster than the dynamics of interest (set by $\Omega V_0 \sim$ tens of MHz). These fast-oscillating terms time-average to zero over any gate duration and are dropped. This is the RWA.

After RWA, the drive Hamiltonian in the rotating frame is:
$$\hat{H}'_d = \frac{\Omega V_0 s(t)}{2} [(-I \cos \delta\omega t + Q \sin \delta\omega t) \hat{\sigma}_x + (I \sin \delta\omega t + Q \cos \delta\omega t) \hat{\sigma}_y]$$

In matrix form using complex exponential representation:
$$\hat{H}'_d = -\frac{1}{2}\frac{C_d}{C_\Sigma}\frac{\hbar}{\sqrt{2Z_0}} V_0 s(t) \begin{pmatrix} 0 & e^{i(\delta\omega t + \phi)} \\ e^{-i(\delta\omega t + \phi)} & 0 \end{pmatrix}$$

\subsubsection*{On-Resonance Simplification}

Setting $\delta\omega = \omega_q - \omega_d = 0$ (drive exactly at qubit frequency), the Hamiltonian reduces to:
$$\hat{H}'_d = -\frac{\Omega V_0 s(t)}{2} (I \hat{\sigma}_x + Q \hat{\sigma}_y)$$

This is a static (non-oscillating) Hamiltonian in the rotating frame—exactly the form needed for universal qubit control:

\begin{itemize}
    \item $\phi = 0$ (pure X drive): $I = 1, Q = 0 \to \hat{H}'_d = -\frac{\Omega V_0 s(t)}{2} \hat{\sigma}_x$ (rotation around X-axis)
    \item $\phi = \pi/2$ (pure Y drive): $I = 0, Q = 1 \to \hat{H}'_d = -\frac{\Omega V_0 s(t)}{2} \hat{\sigma}_y$ (rotation around Y-axis)
    \item Arbitrary $\phi$: $I = \cos\phi, Q = \sin\phi$ (rotation around an axis at angle $\phi$ in the XY plane)
\end{itemize}

\subsection{Rotation Angle and Gate Implementation}

\subsubsection*{The Rotation Angle Integral}

The total rotation angle accumulated by a drive pulse with envelope $s(t)$ and amplitude $V_0$ is:
$$\Theta(t) = -\Omega V_0 \int_0^t s(t') dt'$$

where $\Omega = \frac{C_d}{C_\Sigma}\sqrt{\frac{\hbar}{2Z_0}}$ is a fixed hardware parameter. The integral $\int s(t') dt'$ is the pulse area—determined by the waveform shape and duration. To implement a specific gate, one adjusts:

\begin{itemize}
    \item $V_0$ (pulse amplitude): directly scales $\Theta$
    \item $s(t)$ waveform (shape): common choices are rectangular, Gaussian, or DRAG-shaped envelopes
    \item $t_{\text{pulse}}$ (duration): longer pulse at fixed amplitude gives larger $\Theta$
\end{itemize}

A sequence of gates $(\Theta_k, \Theta_{k-1}, \ldots, \Theta_0)$ is compiled into a sequence of microwave pulses, each with the appropriate amplitude, duration, phase, and shape.

\subsubsection*{Pulse Shaping and DRAG}

Rectangular pulses have broad frequency spectra that can off-resonantly drive the $|1\rangle \to |2\rangle$ transition (separated from $|0\rangle \to |1\rangle$ by the anharmonicity $\alpha \sim -200$ MHz). This causes leakage out of the qubit subspace.

With a Gaussian envelope $s(t)$ on the X channel, the DRAG correction adds:
$$s_Q(t) = -\frac{1}{\alpha} \frac{ds_I(t)}{dt}$$

This doubles the effective gate fidelity for typical transmon anharmonicities, enabling gate errors below $10^{-3}$.

\subsubsection*{Hardware: I/Q Modulation Setup}

In the lab, a microwave local oscillator (LO) at frequency $\omega_{\text{LO}}$ is mixed with two low-frequency signals from an Arbitrary Waveform Generator (AWG) at frequency $\omega_{\text{AWG}}$:
$$\omega_d = \omega_{\text{LO}} + \omega_{\text{AWG}}$$

The AWG generates the $I(t)$ and $Q(t)$ envelopes. An IQ mixer produces the final drive signal at $\omega_d$. By generating multiple AWG tones ($\omega_{\text{AWG1}}, \omega_{\text{AWG2}}, \ldots$) and mixing them with the same LO, one can simultaneously address multiple qubits or readout resonators at different frequencies—a key requirement for scaled-up processors.

\subsection{The Virtual Z Gate}

\subsubsection*{Z Rotations and Phase Shifting}

A physical Z gate rotates the Bloch vector around the Z-axis by angle $\phi$. In the lab frame, this would require either:
\begin{enumerate}
    \item Detuning the drive from $\omega_q$ for a controlled time (allowing free precession to accumulate Z rotation), or
    \item Applying a flux pulse to shift $\omega_q$ temporarily.
\end{enumerate}

Both approaches introduce extra time and potential errors. There is, however, a much simpler and error-free alternative: the virtual Z gate.

\subsubsection*{Why It Works}

Recall that the rotation axis in the XY plane is set by the phase $\phi$ of the drive ($I = \cos\phi, Q = \sin\phi$). In the rotating frame, applying a Z rotation by angle $\phi$ before an X gate is equivalent to relabeling the X and Y axes—i.e., shifting all subsequent phase references by $\phi$. Mathematically:
$$Z(\phi) \cdot X(\theta) \cdot Z^\dagger(\phi) = R_{\text{rotated axis}}(\theta)$$

So instead of explicitly applying $Z(\phi)$, we update the I/Q components for all future pulses:
$$I' = I \cos\phi - Q \sin\phi, \quad Q' = I \sin\phi + Q \cos\phi$$

This is a rotation of the $(I, Q)$ vector by angle $\phi$ on the complex plane—a trivial AWG software update.

\subsubsection*{Benefits and Scope}

Virtual Z gates are now the standard method for implementing Z-axis rotations in superconducting quantum processors:

\begin{itemize}
    \item \textbf{Zero gate time:} No additional pulse is required, improving circuit depth capacity
    \item \textbf{Zero control error:} The rotation is exact to numerical precision, unaffected by pulse distortions or calibration drift
    \item \textbf{No additional decoherence:} Since no physical interaction with the qubit occurs, $T_1$ and $T_2$ are not impacted
    \item \textbf{Composable:} Any sequence of Z rotations can be accumulated and folded into a single phase update
\end{itemize}

In practice, hardware compilers (e.g., IBM Qiskit Transpiler, Google Cirq) automatically convert all Z gates in a quantum circuit to virtual Z gates at the pulse scheduling stage, then propagate the accumulated phase through all subsequent I/Q pulses.

\newpage
\section{Qubit Characterization}

Qubit characterization refers to the experimental techniques and measurements used to assess the performance and quantum properties of a qubit. Understanding decoherence mechanisms and coherence times is essential for optimizing quantum processor design and operation.

\subsection{Open Quantum Systems and Types of Noise}

In an ideal closed quantum system, the qubit's evolution is entirely governed by its Hamiltonian:
$$|\psi(t)\rangle = e^{-i\hat{H}t/\hbar}|\psi_0\rangle$$

which is perfectly deterministic and reversible. In practice, no qubit is truly isolated. It inevitably interacts with its environment—substrate phonons, surface spins, electromagnetic fluctuations in control wiring, thermal photons in the resonator—and this coupling introduces noise.

Noise in quantum systems falls into two broad categories with fundamentally different consequences and remedies.

\subsubsection*{Systematic Noise}

A typical example: a microwave pulse intended to apply an $X(\pi)$ rotation (a NOT gate) is miscalibrated, so it actually applies $X(\pi + \epsilon)$ for some small constant $\epsilon$. Every subsequent application over-rotates by the same $\epsilon$. Over $N$ gates, the error accumulates to $N\epsilon$—detectable and correctable by recalibration. Systematic errors can be diagnosed with randomized benchmarking (which amplifies coherent errors) and corrected via pulse calibration routines (e.g., DRAG parameter sweeps, AllXY sequences).

\subsubsection*{Stochastic Noise}

Common sources of stochastic noise in superconducting qubits include:

\begin{itemize}
    \item \textbf{Thermal Johnson-Nyquist noise:} Voltage noise in $50~\Omega$ control lines with spectral density $S_V(f) = 4k_B T R$, driving unwanted qubit transitions
    \item \textbf{Charge noise:} Random fluctuation of offset charges on the qubit island, causing $\omega_{01}$ to fluctuate—dominant in charge qubits, strongly suppressed in transmons ($E_J/E_C \gg 1$)
    \item \textbf{Flux noise:} Random magnetic flux threading the SQUID loop, causing $\omega_{01}$ to fluctuate—dominant in flux-tunable transmons near flux-sensitive operating points
    \item \textbf{Two-level systems (TLS):} Microscopic defects at metal-oxide interfaces that fluctuate between two configurations, coupling to the qubit electric field and causing both relaxation and dephasing
    \item \textbf{Quasiparticle poisoning:} High-energy particles breaking Cooper pairs and injecting non-equilibrium quasiparticles that transiently shift qubit frequencies
\end{itemize}

\subsection{The Bloch Sphere, Density Matrix, and Open Systems}

\subsubsection*{Bloch Sphere Parametrization}

Any pure state of a qubit can be written as:
$$|\psi\rangle = \cos\frac{\theta}{2}|0\rangle + e^{i\phi}\sin\frac{\theta}{2}|1\rangle$$

where $\theta \in [0, \pi]$ is the polar angle from the north pole ($|0\rangle$) and $\phi \in [0, 2\pi)$ is the azimuthal angle. The corresponding Bloch vector is:
$$\vec{r} = (\sin\theta\cos\phi, \sin\theta\sin\phi, \cos\theta)$$

Key geometric correspondences:
\begin{itemize}
    \item $|0\rangle$: north pole $(\theta=0, \vec{r} = \hat{z})$
    \item $|1\rangle$: south pole $(\theta=\pi, \vec{r} = -\hat{z})$
    \item $|+\rangle = \frac{1}{\sqrt{2}}(|0\rangle+|1\rangle)$: $+x$ axis $(\theta=\pi/2, \phi=0)$
    \item $|+i\rangle = \frac{1}{\sqrt{2}}(|0\rangle+i|1\rangle)$: $+y$ axis $(\theta=\pi/2, \phi=\pi/2)$
\end{itemize}

In the lab frame, a superposition state precesses around the z-axis at the qubit frequency $\omega_{01} \sim 2\pi \times 5$ GHz. In the rotating frame (co-rotating at $\omega_{01}$), this precession is eliminated and the Bloch vector appears stationary—the natural frame for visualizing gate operations and noise effects.

\subsubsection*{Density Matrix and Open Systems}

For an open system coupled to an environment, the state is described by a density matrix:
$$\hat{\rho} = \begin{pmatrix} \rho_{00} & \rho_{01} \\ \rho_{10} & \rho_{11} \end{pmatrix} = \frac{1}{2}(I + \vec{a} \cdot \vec{\sigma})$$

The diagonal elements $\rho_{00}$ and $\rho_{11}$ are populations (probabilities). The off-diagonal elements $\rho_{01} = \rho_{10}^*$ are coherences—they encode the phase relationship between $|0\rangle$ and $|1\rangle$.

Noise has distinct effects on populations and coherences:
\begin{itemize}
    \item \textbf{Longitudinal relaxation ($T_1$):} Drives $\rho_{11} \to 0$, $\rho_{00} \to 1$ (population decays to ground state)
    \item \textbf{Pure dephasing ($T_\phi$):} Drives $\rho_{01}, \rho_{10} \to 0$ while leaving populations unchanged (coherences lost, populations intact)
\end{itemize}

The Lindblad master equation generalizes the Schrödinger equation to open systems, adding dissipation terms that model these noise channels.

\subsection{Longitudinal Relaxation: $T_1$}

\subsubsection*{Physical Mechanism}

Longitudinal relaxation is the spontaneous decay of the excited state $|1\rangle$ to $|0\rangle$ via energy transfer to the environment. It is caused by transverse noise fluctuating at the qubit transition frequency $\omega_{01}$ (Fermi's golden rule):
$$\Gamma_{1\downarrow} \propto S_x(\omega_{01}) + S_y(\omega_{01})$$

where $S_x(\omega)$ and $S_y(\omega)$ are power spectral densities of transverse noise.

\subsubsection*{Temperature Dependence}

The total relaxation rate combines spontaneous emission ($\Gamma_{1\downarrow}$) and thermal excitation ($\Gamma_{1\uparrow}$):
$$\Gamma_1 = \Gamma_{1\downarrow} + \Gamma_{1\uparrow}, \quad \Gamma_{1\uparrow} = \Gamma_{1\downarrow} \cdot e^{-\hbar\omega_{01}/(k_B T)}$$

At $\omega_{01}/2\pi = 5$ GHz and $T = 20$ mK:
$$\frac{\hbar\omega_{01}}{k_B T} \approx 12 \quad \Rightarrow \quad \Gamma_{1\uparrow} / \Gamma_{1\downarrow} \approx e^{-12} \approx 10^{-5}$$

Thermal excitation is negligible—the qubit sees essentially zero thermal photons, which is why dilution refrigerators operating near 20 mK are essential.

\subsubsection*{Lindblad Master Equation for $T_1$}

The density matrix evolution under longitudinal relaxation is:
$$\frac{d\hat{\rho}}{dt} = \Gamma_{1\downarrow} \left(\hat{\sigma}_- \hat{\rho} \hat{\sigma}_+ - \frac{1}{2}\{\hat{\sigma}_+\hat{\sigma}_-, \hat{\rho}\}\right)$$

Solving component-wise:
$$\rho_{11}(t) = \rho_{11}(0) e^{-t/T_1}$$
$$\rho_{01}(t) = \rho_{01}(0) e^{-t/(2T_1)}$$

The population decays with rate $1/T_1$, while coherences decay with rate $1/(2T_1)$—a key observation that separates relaxation's contribution to decoherence from its effect on populations.

\subsection{Pure Dephasing: $T_\phi$}

\subsubsection*{Physical Mechanism}

Pure dephasing does not change populations ($\rho_{00}$ and $\rho_{11}$ remain constant) but destroys phase coherence of superposition states ($\rho_{01} \to 0$). It is caused by longitudinal noise that shifts the qubit frequency randomly:
$$H_{\text{noise}}(t) = \frac{\delta\omega(t)}{2}\hat{\sigma}_z$$

In the rotating frame, this noise causes the Bloch vector to precess at a random rate $\delta\omega(t)$ around the z-axis. Different experimental realizations accumulate different phases, and ensemble averaging washes out transverse components:
$$\rho_{01}(t) = \rho_{01}(0) \left\langle e^{i\int_0^t \delta\omega(t')dt'} \right\rangle_{\text{ensemble}} \to 0$$

\subsubsection*{Noise Sources}

Pure dephasing is dominated by low-frequency (quasi-static) noise sources:
\begin{itemize}
    \item $1/f$ charge noise from interface trap states in oxide layers
    \item Flux noise from trapped flux and fluctuating magnetic spins
    \item Telegraph noise from material defects switching between two states
\end{itemize}

The accumulated phase error under static noise is:
$$\delta\phi(t) = \int_0^t \delta\omega(t') dt'$$

For slow or quasi-static noise, this becomes a runaway phase error unless refocused via spin echo techniques.

\subsection{Total Decoherence: $T_2$ and $T_2^*$}

\subsubsection*{Relationship Between Timescales}

The transverse relaxation time captures total phase coherence loss:
$$\frac{1}{T_2} = \frac{1}{2T_1} + \frac{1}{T_\phi}$$

The factor $1/(2T_1)$ arises because a single energy relaxation event irreversibly collapses the superposition and randomizes the phase. The measured relaxation time $T_2^*$ (observed in free induction decay) includes all dephasing sources:
$$\frac{1}{T_2^*} = \frac{1}{2T_1} + \frac{1}{T_\phi} + \text{(quasi-static noise)}$$

The physical picture: a superposition state on the equator of the Bloch sphere loses its transverse component due to:
\begin{itemize}
    \item Energy relaxation: the Bloch vector collapses to a pole, erasing all phase information
    \item Pure dephasing: the Bloch vector spreads azimuthally due to random phase kicks
    \item Quasi-static fluctuations: different shots experience different effective frequencies
\end{itemize}

The maximum possible $T_2^*$ is limited by $T_1$:
$$T_2^* \leq 2T_1 \quad \text{(equality when } T_\phi \to \infty \text{)}$$

For transmons, $T_2^* \approx 2T_1$ is often approached, meaning they are primarily $T_1$-limited. In other qubit technologies (spin qubits, charge qubits), pure dephasing dominates and $T_2 \ll 2T_1$.

\subsection{Measurement Protocols for Decoherence Timescales}

\subsubsection*{$T_1$ Measurement (Inversion Recovery)}

\paragraph{Pulse Sequence:}
\begin{enumerate}
    \item Apply $X_\pi$ pulse to place qubit in state $|1\rangle$
    \item Wait variable time $\tau$
    \item Measure qubit in the computational basis
    \item Repeat 1000--10000 times for each $\tau$ to obtain average population
    \item Fit decay curve to exponential: $P_1(\tau) = A e^{-\tau/T_1} + B$
\end{enumerate}

The fitting parameters $A$ and $B$ account for initial state preparation fidelity and measurement offset. Typical transmon values: $T_1 \sim 200$--$500~\mu\text{s}$.

\subsubsection*{$T_2^*$ Measurement (Ramsey Interferometry)}

The Ramsey experiment measures total transverse coherence decay including quasi-static noise.

\paragraph{Pulse Sequence:}
\begin{enumerate}
    \item Apply $X_{\pi/2}$ pulse (on-resonance) to place qubit on the equator
    \item Intentionally detune drive by $\delta\omega$ (e.g., 1 MHz off-resonance)
    \item Wait variable time $\tau$; the Bloch vector precesses around the effective Z-axis at rate $\delta\omega$
    \item Apply second $X_{\pi/2}$ pulse with phase aligned to the first
    \item Measure qubit state in computational basis
\end{enumerate}

The measurement signal oscillates and decays:
$$P(|1\rangle) = \frac{1}{2}\left(1 + \cos(\delta\omega \tau) e^{-\tau/T_2^*}\right)$$

The fitted decay time $T_2^*$ includes contributions from intrinsic relaxation and dephasing plus quasi-static frequency fluctuations. Typical transmon values: $T_2^* \sim 10$--$100~\mu\text{s}$.

\subsubsection*{$T_2$ Measurement (Spin Echo / Hahn Echo)}

Ramsey measurements are sensitive to quasi-static (slow-varying) noise. The spin echo technique refocuses quasi-static dephasing by inserting a $\pi$-pulse at the midpoint of the delay.

\paragraph{Pulse Sequence:}
\begin{enumerate}
    \item Apply $X_{\pi/2}$ pulse (on-resonance, $\delta\omega = 0$) to place qubit in superposition
    \item Wait time $\tau/2$
    \item Apply $X_\pi$ pulse (180° refocusing flip)
    \item Wait time $\tau/2$
    \item Apply second $X_{\pi/2}$ pulse, phase-aligned to first
    \item Measure
\end{enumerate}

\paragraph{Refocusing Mechanism:} The central $\pi$-pulse acts as a time-reversal for slow noise. If the qubit sees a constant frequency offset $\delta$ during $[0, \tau/2]$, it accumulates phase $+\delta\tau/2$. After the $\pi$-flip, the same offset causes it to accumulate $-\delta\tau/2$, so quasi-static noise cancels. Only noise fluctuating faster than $\tau/2$ survives.

\paragraph{Measurement Result:} The echo decay is:
$$P(|1\rangle) = \frac{1}{2}(1 + e^{-\tau/T_2})$$

Because slow noise is refocused, $T_2$ (spin echo time) is always $\geq T_2^*$ (Ramsey time). The relationship chain is:
$$T_2^* \leq T_2 \leq 2T_1$$

This hierarchy provides noise spectroscopy:
\begin{itemize}
    \item Gap between $T_2^*$ and $T_2$: indicates quasi-static noise contribution
    \item Gap between $T_2$ and $2T_1$: indicates non-refocusable intermediate-frequency noise
    \item If $T_2 \approx 2T_1$: qubit is primarily $T_1$-limited (ideal for transmons)
\end{itemize}

Typical transmon values: $T_2 \sim 100$--$300~\mu\text{s}$.

\subsection{Summary Table: Decoherence Timescales and Measurement Methods}

\begin{center}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Parameter} & \textbf{$T_1$} & \textbf{$T_2^*$ (Ramsey)} & \textbf{$T_2$ (Echo)} \\
\hline
\textbf{Physical Process} & Energy relaxation & Total dephasing & Refocused dephasing \\
\hline
\textbf{Noise Type} & Transverse & Longitudinal + quasi-static & Longitudinal (fast) \\
\hline
\textbf{Mechanism} & $|1\rangle \to |0\rangle$ decay & Random phase shifts & Phase randomization \\
\hline
\textbf{Bloch Effect} & Z-axis collapse & XY-plane decay & XY-plane decay \\
\hline
\textbf{Measurement} & Inversion recovery & Free induction decay & Spin echo \\
\hline
\textbf{Decay Formula} & Exponential & Oscillation + decay & Exponential decay \\
\hline
\textbf{Typical Value} & 200--500 $\mu$s & 10--100 $\mu$s & 100--300 $\mu$s \\
\hline
\textbf{Relationship} & $T_2 \leq 2T_1$ & $T_2^* \leq T_2$ & $T_2 \geq T_2^*$ \\
\hline
\end{tabular}
\end{center}

\subsection{Performance Metrics and Quantum Advantage}

For scalable quantum computing, all decoherence timescales must be much longer than gate times. Modern superconducting qubits achieve gate times of 10--100 ns. The quality factor is:
$$Q_{\text{gate}} = \frac{T_2}{t_{\text{gate}}}$$

For fault-tolerant quantum computation under surface codes:
$$Q_{\text{gate}} > 1000$$

This requires $T_2 > 1~\mu\text{s}$ when $t_{\text{gate}} \sim 1$ ns, or $T_2 > 100~\mu\text{s}$ when $t_{\text{gate}} \sim 100$ ns. State-of-the-art transmon qubits achieve $Q_{\text{gate}} \sim 1000$--$3000$, approaching the threshold for error correction.


\end{document}